Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\ (New project paper)

### 2016-03-31 pubs/tfp16/paper.tex
some edits and thoughts

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Non-strictness
constrains the degree of potential parallelism, that is, we are not exploring
speculative evaluation.  A secondary goal is to experiment with mechanisms
for transparent fault tolerance.
\\
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
  \item We are reporting work in progress.  At the time of writing we have a
    complete serial implementation, a primitive proof-of-concept parallel
    implementation, and are currently developing a more realistic parallel
    runtime.  We expect to have preliminary results from our new parallel
    implementation by the time of the workshop.  The fault-tolerance mechanisms
    remain future work.

@@

  \item Hopefully in the spirit of a student-friendly workshop, for this draft
    version we have chosen a more informal, conversational style of
    presentation than necessarily appropriate for a final paper, have
    highlighted the major student contributions and involvement, and included
    perhaps gratuitous detail about process.
  \end{enumerate}

@@

\subsection{The trend: strict but still pure functional programming}
There is a small but growing understanding that \emph{pure functional
  semantics}, in some form, is essential to reining in the complexity of these
ever-evolving scientific codes.  Such semantics are expressed in a number
of ways as highlighted following.

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all; indeed, Fortran
was designed to make this easy via \texttt{COMMON} blocks---aggregates of
global variables---which further obfuscate program meaning by allowing
arbitrary naming of a block's variables at a subroutine/function granularity,
and requiring that the entire block be brought into scope.  It is almost
ironic, then, that of the mainstream high-performance scientific programming
languages (Fortran, C, and C++), Fortran~95 was the first to introduce the
\texttt{pure} function qualifier, and with enforcement by the compiler.
Otherwise it is entirely up to the programmer to make functions composable and
thread safe.  Here at LANL this approach is being used to good effect to clean
up some legacy code bases.

\subsubsection{Obey constraints prescribed by a parallel runtime system.}
There are a number of parallel runtime systems that encourage (but cannot
enforce) a pure functional style.  We will highlight Stanford's Legion
runtime as an example~\cite{Bauer12}.

In Legion, non-function-local data, or \emph{regions}, are requested from the
runtime system.  Regions may be shared among \emph{tasks} (essentially C++
functions), and each task is prescribed access privileges (read, write,
read/write, etc.)  and coherency requirements for each region to which it will
have access.  Other than registered access to regions, tasks must not access
non-constant global data.  In serial execution these access requirements are
notionally superfluous, but in parallel execution the runtime can dynamically
calculate the data dependency graph and relax the programmatically defined
serial ordering of task execution.  Very high parallel efficiency has been
demonstrated on extremely large computing systems~\cite{Bauer14}.  Here at
LANL the Legion system is being strongly considered for a new large-scale C++
code effort.\footnote{It seems significant that the PI of the Legion
  project was previously a co-developer of a strict, arguably pure functional
  language implementation~\cite{AikenFL,FLreport89}, in collaboration with
John Backus as a refinement of his FP language~\cite{Backus:1978}.}

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and non-strictness or laziness is more of a
nuisance--even a serious hindrance--than a convenience, much less ever
algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84}
or Johnsson's general attribute grammars~\cite{Johnsson87}.

@@

Much anecdotal evidence suggests that many others have similar difficulties
with lazy (or non-strict) by default semantics, with the most telling recent
concrete evidence being the push for, and realization of, the \texttt{Strict}
and \texttt{StrictData} language extensions newly available in GHC 8.0.
Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practioners are by and large not computer scientists}.
The concept of a space leak, for example, caused by other than a missing explicit
deallocation statement (C \emph{free}, etc.) is difficult to understand and not
something they particularly want to try to understand.  Hence, we eschew explicit
strictness annotations for varying uniform degrees of strictness.

@@

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various point in time sufficient program state is dumped to backing store such
that the program, in case of a fault, can restart from the latest stored
state.  Most implementations are explicit, though there has been some success
with transparent (invisible to the programmer) systems.  Wiring
checkpoint/restart into a large-scale application is not only burdensome and
error-prone, it can severely warp the engineering of a large-scale application
because it must be designed such that essential state can be readily captured.
And now, it is widely believed that the scalability of checkpoint/restart may
be reaching its end.

In the imperative world side-effect-free tasks could be safely restarted after
failure, needing only their well-defined inputs and not arbitrary access to
mutable global state.  The Legion system is being augmented with this
capability, and in such a way that it will be largely transparent to the programmer.
Ericsson's quasi-functional language Erlang and its runtime system was
designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been
recognized that the pure parts of a functional program (excluding, e.g., the
I/O monad in Haskell) could be safely restarted.  The recent work of
[Trinder's student] impressively demonstrates this for an augmented GHC/runtime
on a large-scale distributed-memory system.

We claim, then, that there is a trend, in HPC at least, towards
\emph{strict-by-default pure functional programming}, whether as imposed by
programmer discipline, a parallel runtime system, or the language itself
(e.g.\ strict Haskell).  Second, we claim that there is a trend in
high-performance computing, and by transitivity functional programming,
towards transparent fault tolerance.

\subsection{Strictness and Parallelism}

It is a trivial observation that given strict functional semantics, wherein
function arguments may be safely evaluated before function evaluation, it is
safe to evaluate the arguments, and the function, in parallel, and that this
principle applies recursively in the (dynamic) expression tree (or graph).

\section{Project and Goals}

Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict, and secondarily, with mechanisms
for transparent fault tolerance.  Light-weight has several meanings or
implications.  First and foremost it means that \emph{we are not attempting to
  compete with GHC in any respect}---succumbing to the temptation of
recreating GHC language, type system, and runtime features, or significant
code optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that GHC currently
  represents, or put another way, one second for each hour, respectively.}

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\end{itemize}

@@

\subsection{Varying degrees of strictness}

First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term \emph{lazy} and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense, unless stated otherwise.
Similarly, when referring to language semantics we refer by default to the
operational rather than denotational semantics.

@@

\section{Design and Implementation}

Our wish list has much in common with the GHC Haskell implementation, and
indeed GHC is part of our master plan.  In brief, the phases of the GHC
compiler are 1) parsing and type checking; 2) transformation to the GHC
\emph{Core} intermediate representation; 3) analysis and transformation on
Core; 4) transformation of Core to the \emph{STG} intermediate representation;
5) transformation of STG to machine code via one of various imperative
representations (Cmm, LLVM, C).

In short, our plan is to use GHC as our front-end to generate its intermediate
form, STG~\cite{PJ-stockhardware}, then escape to our system from there.
There are various possible ways to accomplish the former: using GHC's option
to dump STG as text, hacking GHC itself and, optionally, using GHC as a
library~\cite{ghc-as-library}.  An alternative would be to start with GHC Core.

\subsection{STG Language}
\setlength{\tabcolsep}{5pt}

Our incarnation of STG, as is our presentation in Figure~\ref{fig:STGsyntax} which
is meant to evoke both an abstract and concrete syntax, are very similar to
that given in Marlow and Peyton Jones' \emph{fast curry}
paper~\cite{Marlow:2004}, to which we refer so many times that we will
subsequently refer to it as simply \emph{EA}.\footnote{\emph{Eval/Apply} from
  the title because \emph{FC} likely has another connotation to the reader.}

\begin{figure}
%\centering
\begin{centering}
\footnotesize % tiny scriptsize footnotesize small
\begin{tabular}{r r c l l}

@@

Partial application (\emph{PAP}) and black hole (\emph{BLACKHOLE}), i.e., a thunk being
evaluated, are only runtime entities and not actually part of our STG language.

The semantics of STG is what one would expect for a higher-order, polymorphic,
pure functional language, with one exception for a Haskell programmer familiar
with unboxed types:  \texttt{case} is strict in the scrutinee, so
\texttt{case} is similar to \texttt{pseq} (as opposed to \texttt{seq}) in Haskell,
that is, the scrutinee is evaluated first.

In STG---one of GHC's
intermediate representations---the distinction is clear: heap objects are
explicitly constructed by \texttt{let} constructs in which constructor names
are given explicitly (not aliased); constructor names are not functions.

A distinctive feature of STG is that heap allocation is explicit; this is exactly
what \texttt{let} bindings do.  As such unboxed values cannot be let-bound; instead
unboxed values can be bound with \texttt{case}:
\begin{verbatim}
    case e of x# {
      _ -> e
    }
\end{verbatim}
where the \texttt{\#} suffix is merely a convention to signify an unboxed type.
Deferred evaluation is made explicit:  a deferred expression evaluation is
encoded as a thunk.

\subsection{Degrees of strictness, take two}

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG.  Our partial description is based on that given in \emph{EA}.

@@

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
(partial application of a \emph{FUN}) is determined---in the case of
\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
arguments previously provided.  If the arity is greater than two then a new
\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
the new \emph{PAP} returned.  If the arity is equal to two then the underlying
\emph{FUN} is tail-called with any previous argument (in case of \emph{PAP})
and new arguments \texttt{x} and \texttt{y}.  If the arity is less than two
(it must be at least one, so one in this case) the underlying \emph{FUN} is
called-with-return with any previous arguments (in case of \emph{PAP}) and new
argument \texttt{x}, then the object returned applied to remaining argument
\texttt{y}.

Given application \texttt{f x y} when might \texttt{x} and \texttt{y} be
evaluated?  In the non-strict semantics evaluation is on demand, after
\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one of
those arguments forced in subsequent evaluation (assuming that they weren't
forced in the evaluation of \texttt{f} itself).

In a strict semantics one can come up with various evaluation orders, but
excluding the absurd, as far as distinguishing termination properties goes
these can be distilled down to two possibilities, namely whether on
construction of a \emph{PAP} the (new) arguments are forced or not.  (An
absurd strategy might force every second argument.)  These are in fact
distinguishable because (in Haskell) the first argument to \texttt{seq} might
evaluate to a \emph{PAP}.

@@

\subsection{STG to C}

The back-end is written in Haskell and generates proper tail-calling C code;
the runtime system is written in C/C++.  The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which, of course, may be different than those for non-strict
semantics.  Sharing is preserved.

\subsubsection{Proper tail calls.}
While performance is not a high priority, we attain proper tail calling instead of
using the trampoline used in earlier versions of GHC~\cite{PJ-stockhardware}.
Because we maintain our own data and control (continuation) stack all
generated C functions take no explicit arguments and have return type void,
and with a single exception in the non-strict case are invoked in tail call
positions, at non-zero optimization levels both gcc and Clang/LLVM generate
jumps instead of calls for C calls in tail call positions.  When using gcc
we also use a fixed set of machine registers for passing/returning values.

The exceptions to tail calling are for convenience.  Similar to what is
described in \emph{EA} we have a generic variadic \emph{stgApply} function
that is used for applications of non-known functions, and known functions with
a deficit or excess number of arguments.  The \emph{stgApply} function takes
as arguments the first element of the application (\emph{FUN}, \emph{PAP},
\emph{THUNK}, or \emph{BLACKHOLE}) and its arguments, and if a \emph{THUNK},
evaluates it with a call (rather than jump) to yield a \emph{FUN} or
\emph{PAP}.  This allows us to implement a single \emph{stgApply} function as
a C function taking a contiguous block of machine words (unboxed values
or pointers) on the explicit stack.  The implementation of \emph{stgApply}
as described in \emph{EA} makes the same concession regarding
call-with-return, but the scheme described there uses an auto-generated family
of \emph{stgApply} functions parameterized by number and type (unboxed or
pointer) of arguments.  We used this scheme initially as well, but because the
parameter block contains a bitmap describing which arguments are boxed/unboxed
for the benefit of the garbage collector, this elaboration is unnecessary.

For \emph{stgApply} in the non-strict semantics an alternative to
call-with-return would be a pair of functions \emph{stgApply1} and
\emph{stgApply2}.  In one possible variation, \emph{stgApply1} would push a
continuation with \emph{stgApply2} as the return address, then invoke the
first element of the application.  Call-with-return becomes more convenient
when considering the various forms of strictness: when evaluating function or
constructor arguments it is simpler, and arguably more efficient in terms of
time (but not space in general) to ``evaluate in place'' than implement the
continuation style to the logical limit.

\subsection{Mini-Haskell}
As mentioned, a current near-term goal is to use GHC as our front-end to
generate STG as input to our system.  In the meantime, as a stop-gap measure
to make the generation of test cases easier, we have implemented a
\emph{mini-Haskell} front-end that supports a subset of standard Haskell
syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote an
unboxed literal value.  Our extension (to STG as well)
is the provision of the optional \texttt{unboxed} keyword in datatype definitions
to allow user-defined unboxed data types, e.g.,
\begin{verbatim}
  data unboxed Bool# = False# | True# .
\end{verbatim}

Writing STG in the concrete syntax conceived for our project is tedious and
prone to errors.  Compared to Haskell, the eventual source language for the
compiler, STG is relatively verbose, requiring explicit block and expression
delimiters, and classification of heap objects. Though its general structure
is similar to primitive Haskell, it lacks many of the conveniences that make
Haskell a pleasure to program in.  The motivation for a more expressive
language for testing our compiler and runtime system in their nascent stages
is then clear.  This language, mini-Haskell, was progressively ``grown'' from
the STG syntax towards a subset of Haskell itself*, removing much of the
frustrating syntactic cruft and adding some of the core features that are
often taken for granted in Haskell (and other languages).

@@

At the time of writing we have a complete serial implementation, including
\begin{itemize}
\item STG (inc. ADT w/unboxed) parser
\item type inferencer
\item various analysis and transformation passes
\item code generator
\item runtime system including a garbage collector
\end{itemize}

@@

Trinder--speculative parallelism

Trinder's paper (or just web site?) on fault tolerance.

Trinder's student's thesis on fault tolerance.

>>>>

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Non-strictness
constrains the degree of potential parallelism, that is, we are not exploring
speculative evaluation.  A secondary goal is to experiment with mechanisms
for transparent fault tolerance.
\\
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
  \item We are reporting work in progress.  At the time of writing we have a
    complete serial implementation, a primitive proof-of-concept parallel
    implementation, and are currently developing a more realistic parallel
    runtime.  We expect to have preliminary results from our new parallel
    implementation by the time of the symposium.  The fault-tolerance mechanisms
    remain future work.

@@

  \item Hopefully in the spirit of a student-friendly symposium, for this
    draft version we have chosen a more informal, conversational style of
    presentation than necessarily appropriate for a final paper, have
    highlighted student involvement and its influence on our approach, and
    included perhaps gratuitous detail about process.
  \end{enumerate}

@@

\subsection{The trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new, as evidenced
by the Workshop on Functional High Performance Computing, held in conjuction
with the International Conference on Functional Programming, in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of these ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all; indeed, Fortran
was designed to make this easy via \texttt{COMMON} blocks---aggregates of
global variables---which further obfuscate program meaning by allowing
arbitrary naming of a block's variables at a subroutine/function granularity,
and requiring that the entire block be brought into scope.  It is almost
ironic, then, that of the mainstream high-performance scientific programming
languages (Fortran, C, and C++), Fortran~95 was the first to introduce the
\texttt{pure} function qualifier, and with enforcement by the compiler.
Otherwise it is entirely up to the programmer to make functions composable and
thread safe.  Here at LANL the use of pure functions (among other techniques
for reducing global data dependencies) is being used to good effect to clean
up some legacy code bases.

\subsubsection{Obey constraints prescribed by a parallel runtime system.}
There are a number of parallel runtime systems that encourage (but cannot
enforce) a pure functional style.  We highlight Stanford's Legion
runtime as an example~\cite{Bauer12}.

Using Legion, all non-function-local variables, or \emph{regions}, must
requested from the runtime system.  Regions may be shared among \emph{tasks}
(essentially C++ functions), and each task is prescribed access privileges
(read, write, read/write, etc.) and coherency requirements for each region to
which it will have access.  Other than registered access to regions, tasks
must not access non-constant global data.  In serial execution these access
requirements are notionally superfluous, but in parallel execution the runtime
can dynamically calculate the data dependency graph and relax the
programmatically defined serial ordering of task execution.  Very high
parallel efficiency has been demonstrated on extremely large computing
systems~\cite{Bauer14}.  Here at LANL the Legion system is being strongly
considered for a new large-scale C++ code effort.\footnote{It seems
  significant that the PI of the Legion project was previously a co-developer
  of a strict, arguably pure functional language
  implementation~\cite{AikenFL,FLreport89}, in collaboration with John Backus
  as a refinement of his FP language~\cite{Backus:1978}.}

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and non-strictness or laziness is more of a
nuisance--even a serious hindrance--than a convenience, much less ever
algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
of \emph{tying the knot.}~\cite{tying-the-knot}.

@@

Much anecdotal evidence suggests that many others have similar difficulties
with semantics that are lazy (or non-strict) by default, with the most telling
recent concrete evidence being the push for, and realization of, the
\texttt{Strict} and \texttt{StrictData} language extensions newly available in
GHC 8.0.  Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practioners are by and large not computer
  scientists}, they are domain scientists (physicists, chemists, etc.) or
computational scientists specialized in numerical or numerically-oriented
algorithmic methods (e.g., Lagrangian, Eulerian, adaptive mesh refinement).
The concept of a space leak, for example, caused by other than a missing
explicit deallocation statement (C \emph{free}, etc.) is difficult to
understand and not something they particularly want to try to understand.
Hence, we eschew explicit strictness annotations for varying uniform degrees
of strictness.

@@

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various point in time sufficient program state is dumped to backing store such
that the program, in case of a fault, can be (manually) restarted from the
latest stored state.  Most implementations are explicit, though there has been
some success with transparent (invisible to the programmer) systems.  Wiring
checkpoint/restart into a large-scale application is not only burdensome and
error-prone, it can severely warp the engineering of a large-scale application
because it must be designed such that essential state can be readily captured.
And now, it is widely believed that the scalability of checkpoint/restart may
be reaching its end.

In the imperative world side-effect-free tasks could be safely restarted after
failure, needing only their well-defined inputs and not arbitrary access to
mutable global state.  The Legion system is being augmented with this
capability, and in such a way that it will be largely transparent to the
programmer.  Ericsson's quasi-functional language Erlang and its runtime
system was designed for fault tolerance.~\cite{Cesarini:2009}.  It has long
been recognized that the pure parts of a functional program (excluding, e.g.,
the I/O monad in Haskell) could be safely restarted.  The recent work of
Steward impressively demonstrates this on a large-scale distributed-memory
system~\cite{Stewart:2013}.

We claim, then, that there is a trend, in HPC at least, towards
\emph{strict-by-default pure functional programming}, whether as imposed by
programmer discipline, a parallel runtime system, or the language itself
(e.g.\ strict Haskell).  Second, we claim that there is a trend in
high-performance computing, and by transitivity in functional programming,
towards transparent fault tolerance.

\subsection{Strictness and Parallelism}
It is a trivial observation that given strict functional semantics, wherein
function arguments may be safely evaluated before function evaluation, it is
safe to evaluate the arguments, and the function, in parallel, and that this
principle applies recursively in the (dynamic) expression tree (or graph).

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict, and secondarily, with mechanisms
for transparent fault tolerance.  Light-weight has several meanings or
implications.  First and foremost it means that \emph{we are not attempting to
  compete with GHC in any respect}---succumbing to the temptation of
recreating GHC language, type system, and runtime features, or significant
code optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that GHC currently
  represents, or put another way, one second for each hour, respectively.}
For another, it means that we are not undertaking significant morphing of GHC
itself:  because this is in large part a student project it must be
kept as simple as reasonably possible in various respects.

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline.
\end{itemize}

@@

\subsection{Varying degrees of strictness}
First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term \emph{lazy} and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense, unless stated otherwise.
Similarly, when referring to language semantics we refer by default to the
operational rather than denotational semantics.

@@

\section{Design and Implementation}
Our wish list has much in common with the GHC Haskell implementation, and
indeed GHC is part of our master plan.  In brief, the phases of the GHC
compiler are 1) parsing and type checking; 2) transformation to the GHC
\emph{Core} intermediate representation; 3) analysis and transformation on
Core; 4) transformation of Core to the \emph{STG} intermediate representation;
5) transformation of STG to machine code via one of various imperative
representations (Cmm, LLVM, C).

In short, our plan is to use GHC as our front-end to generate its intermediate
form, STG~\cite{PJ-stockhardware}, then escape to our system from there.
There are various possible ways to accomplish the this: using GHC's option
to dump STG as text, hacking GHC itself and, optionally, using GHC as a
library~\cite{ghc-as-library}.  An alternative would be to start with GHC Core
which we may transition to.

\subsection{STG Language}

Our incarnation of STG, as shown in Figure~\ref{fig:STGsyntax}, is meant to
evoke both an abstract and concrete syntax.  This presentation is adapted from
Marlow and Peyton Jones' \emph{Eval/Apply} paper~\cite{Marlow:2004}, to which
we refer so many times that we will subsequently refer to it as simply
\emph{EA}. Partial application (\emph{PAP}) and black hole (\emph{BLACKHOLE}),
i.e., a thunk being evaluated, are only runtime entities and not actually part
of our STG language.

\setlength{\tabcolsep}{5pt}
\begin{figure}
%\centering
\begin{centering}
\footnotesize % tiny scriptsize footnotesize small
\begin{tabular}{r r c l l}

@@

The semantics of STG is what one would expect for a higher-order, polymorphic,
pure functional language, with a few exceptions for a Haskell programmer
familiar with unboxed types.  First, \texttt{case} is strict in the scrutinee,
so \texttt{case} is similar to \texttt{pseq} (as opposed to \texttt{seq}) in
Haskell, that is, the scrutinee is evaluated first.  Second, constructors are
not functions, rather they are named in the construction (heap allocation) of
a data object.  More generally, the operational notion of allocating a heap
object is made explicit: this is the ``side effect'' of \texttt{let} bindings.
As such a variable of unboxed type cannot be let-bound; instead it can be
bound with \texttt{case}, e.g., \texttt{case e of x\# \{ \_ -> e \}}
%\begin{verbatim}
%    case e of x# {
%      _ -> e
%    }
%\end{verbatim}
where the \texttt{\#} suffix is merely a convention to signify that
\texttt{x\#} has unboxed type.  Finally, deferred evaluation is made explicit:
a deferred expression evaluation is encoded by a thunk.

\subsection{Degrees of strictness, take two}
Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG abstract machine.  Our partial description is based on that given
in \emph{EA}.

@@

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
(partial application of a \emph{FUN}) is determined---in the case of
\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
arguments previously provided.  If the arity is greater than two then a new
\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
the new \emph{PAP} returned.  If the arity is equal to two then the underlying
\emph{FUN} is tail-called (jumped to) with any previous arguments (in case of
\emph{PAP}) and new arguments \texttt{x} and \texttt{y}.  If the arity is less
than two (it must be at least one, so one in this case) the underlying
\emph{FUN} is called-with-return with any previous arguments (in case of
\emph{PAP}) and new argument \texttt{x}, then the object returned applied to
remaining argument \texttt{y}.

Given application \texttt{f x y} when might \texttt{x} and \texttt{y} be
evaluated?  In the non-strict semantics evaluation is on demand, after
\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
both of those arguments forced in subsequent evaluation (assuming that they
weren't forced in the evaluation of \texttt{f} itself).

In a strict semantics one could devise various evaluation orders, but as far
as distinguishing termination properties these can be distilled down to two
reasonable possibilities, namely whether on construction of a \emph{PAP} the
(new) arguments are forced or not.  In other words, the question is whether
application is strict, or actual function call is strict.  These are in fact
distinguishable because (in Haskell) the first argument to \texttt{seq} might
evaluate to a \emph{PAP}.

@@

The other dimension we vary is whether or not constructors are strict.
Other dimensions are possible, for example whether \texttt{let} bindings
(heap allocations) are constructed on demand, but we restrict ourselves
to function/application and constructor strictness.

\subsection{STG to C}
The back-end is written in Haskell and generates proper tail-calling C code;
the runtime system is written in C/C++.  The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which, of course, may be different than those for non-strict
semantics.  Sharing is preserved.

\subsubsection{Proper tail calls.}
While performance is not a high priority, we attain proper tail calling
instead of using the trampoline used in earlier versions of
GHC~\cite{PJ-stockhardware}.  Because we maintain our own data and control
(continuation) stack all generated C functions take no explicit arguments and
have return type \emph{void}, and with a single exception in the non-strict
case are invoked in tail call positions, at non-zero optimization levels both
gcc and Clang/LLVM generate jumps instead of calls for C calls.  When using
gcc we also use a fixed set of machine registers for passing/returning values.

The exceptions to tail calling are for convenience.  Similar to what is
described in \emph{EA} we have a generic variadic \emph{stgApply} function
that is used for applications of non-known functions, and known functions with
a deficit or excess number of the arguments.  The \emph{stgApply} function takes
as arguments the first element of the application (\emph{FUN}, \emph{PAP},
\emph{THUNK}, or \emph{BLACKHOLE}) and its arguments, and if a \emph{THUNK},
evaluates it with a call (rather than jump) to yield a \emph{FUN} or
\emph{PAP}.  This allows us to implement a single \emph{stgApply} function as
a C function taking a contiguous block of machine words (unboxed values
or pointers) on the explicit stack.  The implementation of \emph{stgApply}
as described in \emph{EA} makes the same concession regarding
call-with-return, but the scheme described there uses an auto-generated family
of \emph{stgApply} functions parameterized by number and type (unboxed or
pointer) of arguments.  We used this scheme initially as well, but because the
parameter block contains a bitmap describing which arguments are boxed/unboxed
for the benefit of the garbage collector, this elaboration is unnecessary.
For \emph{direct calls}, in which a known function is applied to a number
of arguments equal to its arity, \emph{stgApply} need not be invoked:
the \emph{FUN} is tail-called directly.

Call-with-return becomes even more convenient when considering the various
forms of strictness: when evaluating function or constructor arguments it is
simpler, and arguably more efficient in terms of time (but not space in
general beca) to ``evaluate in place'' than implement the continuation style
to the logical limit.  We are aware, however, that we may need to eliminate
this use of the native C stack so that we can properly unwind the stack in a
more sophisticated parallel runtime when pruning multiple evaluations of the
same thunk, and we have determined that this is a relatively minor exercise:
one possibility is that \emph{stgApply} takes an additional argument
indicating which of its subsequent arguments have already been forced, and at
each invocation if unforced arguments remain, evaluates the next (possibly)
unevaluated argument and tail calls itself, re-using the stack frame with just
this new argument updated.

\emph{It seems that we can dispose with call-with-return of STG functions,
  including stgApply (as opposed to runtime functions) without too much
  effort.  It also seems like this is needed if we want to be able to unwind
  the stack as described in the ``grey hole'' paper.  I'll try to do this this
  week (3/31/16-4/1/16) to verify the feasibility before changing the
  preceding description (which isn't quite right)}

\subsection{Mini-Haskell}
As mentioned, a current near-term goal is to use GHC as our front-end to
generate STG (or Core) as input to our system.  In the meantime, as a stop-gap
measure to make the generation of test cases easier, we have implemented a
\emph{mini-Haskell} front-end that supports a subset of standard Haskell
syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote
an unboxed literal value.  Our extension (to STG as well) is the provision of
the optional \texttt{unboxed} keyword in datatype definitions to allow
user-defined unboxed data types, e.g.,
\begin{verbatim}
  data unboxed Bool# = False# | True# .
\end{verbatim}

Writing STG in the concrete syntax conceived for our project is tedious and
prone to errors.  Compared to Haskell, the eventual source language for the
compiler, STG is relatively verbose, requiring explicit block and expression
delimiters, and classification of heap objects. Though its general structure
is similar to primitive Haskell, it lacks many of the conveniences that make
Haskell a pleasure to program in.  The motivation for a more expressive
language for testing our compiler and runtime system in their nascent stages
is then clear.  This language, mini-Haskell, was progressively ``grown'' from
the STG syntax towards a subset of Haskell itself, removing much of the
frustrating syntactic cruft and adding some of the core features that are
often taken for granted in Haskell (and other languages).

@@

At the time of writing we have a complete serial implementation, including
\begin{itemize}
\item STG and mini-Haskell (inc. ADT w/unboxed) parsers \emph{(both by students)}
\item mini-Haskell to STG transformer \emph{(by student)}
\item type inferencer
\item various analysis and transformation passes \emph{(all by students)}
\item code generator \emph{(some student contribution)}
\item runtime system including a garbage collector
\item extensive test suite \emph{(mostly by students)}
\end{itemize}

@@
